---
title: "Strava Fitness Analysis"
format: html
editor: visual
embed-resources: true
bibliography: references/references.bib
csl: references/apa.csl
---

# Instructions

```{r, echo=FALSE}
#devtools::install_github("risktoollib/RTLedu")
#RTLedu::fitness()
```

# The Task

1.  Explain how you define fitness.

2.  Tell me about which individual is fitter?

3.  Tell me about which individual has improved more in fitness?

4.  Make a clear and concise recommendation on which individual you would want to coach. There is no `it depends here`, you have to make a decision.

Write a bullet point summary on your findings at the beginning of the document.

You must use a quarto document (.qmd) and render it into `html`.

# Questions

-   Could there be a correlation between personal performance and an individual pursuing their favourite activity?

# What is Fitness?

According to [@Caspersen1985]

# Importing Strava Data

```{r, echo=FALSE}
library(tidyverse)
```

```{r, echo=FALSE}
strava_data <- RTLedu::strava

strava_data
```

```{r, echo=FALSE}
clean_strava_data <- strava_data %>% 
  dplyr::select(- where(~ all(.x == 0))) # The ~ is like lambda in Python and is used to check all of the columns. The .x is for everything is that column that lambda is going to look at

clean_strava_data
```

```{r, echo=FALSE}
clean_strava_data %>% 
  dplyr::filter(Runner == "Hound")
```

```{r, echo=FALSE}

```
