---
title: "Strava Fitness Analysis"
author: Shelby Cyr, Daniel Rabi, and Ali Sachwani
date: March 7, 2026
backgroundcolor: lightblue
#theme: Quartz
format: html
editor: visual
embed-resources: true
bibliography: references/references.bib
csl: references/apa.csl
output-file: index

echo: FALSE
warning: FALSE
message: FALSE
---

```{r}
# Instructions:
#devtools::install_github("risktoollib/RTLedu")
#RTLedu::fitness()
```

```{r}
library(tidyverse)
library(corrplot)
library(lubridate)
library(GGally)
library(slider)
library(patchwork)
```

```{r}
strava_data <- RTLedu::strava
#Initial data import
strava_data
```

```{r}
clean_strava_data <- strava_data %>% 
  dplyr::select(- c(18:27, 29)) %>%  #Removal of empty columns
  dplyr::select(- "Decompression") # Removing this because it's a no for everyone, so we don't to keep it there

clean_strava_data <- clean_strava_data %>% 
  dplyr::mutate(Distance = dplyr::if_else(Activity.Type == "Track Running", Distance / 1000, Distance)) # Converting meters to Km

clean_strava_data
```

```{r}
clean_strava_data <- clean_strava_data %>% 
  dplyr::mutate(Avg.HR = as.numeric(Avg.HR)) %>% 
  dplyr::mutate(Max.HR = as.numeric(Max.HR)) %>%
  dplyr::mutate(Aerobic.TE = as.numeric(Aerobic.TE)) %>%
  dplyr::mutate(Avg.Run.Cadence = as.numeric(Avg.Run.Cadence)) %>%
  dplyr::mutate(Max.Run.Cadence = as.numeric(Max.Run.Cadence)) %>%
  dplyr::mutate(Total.Ascent = as.numeric(Total.Ascent)) %>%
  dplyr::mutate(Total.Descent = as.numeric(Total.Descent)) %>%
  dplyr::mutate(Min.Elevation = as.numeric(Min.Elevation)) %>%
  dplyr::mutate(Max.Elevation = as.numeric(Max.Elevation))
#ensuring data will be easy to work with for future calculations in terms of format.
clean_strava_data
```

```{r}
clean_strava_data <- clean_strava_data %>% 
  dplyr::rename(
    Avg.Cadence = Avg.Run.Cadence,
    Max.Cadence = Max.Run.Cadence,
    Avg.Stride = Avg.Stride.Length
    )

clean_strava_data
```

```{r}
clean_strava_data <- clean_strava_data %>% 
  dplyr::mutate(
    Avg.Pace = period_to_seconds(hms(Avg.Pace)) / 60^2,
    Best.Pace = period_to_seconds(hms(Best.Pace)) / 60^2,
    Best.Lap.Time = period_to_seconds(hms(Best.Lap.Time)) / 60^2
    ) %>% 
  dplyr::mutate(
    TE.score =
    dplyr::if_else (Aerobic.TE >= 0 & Aerobic.TE < 1,0,
          dplyr::if_else (Aerobic.TE >= 1 & Aerobic.TE < 2,2,
              dplyr::if_else (Aerobic.TE >= 2 & Aerobic.TE < 3,4,
                dplyr::if_else (Aerobic.TE >= 3 & Aerobic.TE < 4,6,
                  dplyr::if_else (Aerobic.TE >= 4 & Aerobic.TE < 5,10,0)
    )))))

clean_strava_data
```

```{r}
clean_strava_data <- clean_strava_data %>% 
  dplyr::mutate(Avg.Speed = 60 / Avg.Pace) %>%
  dplyr::mutate(EF = (Avg.Speed / Avg.HR))

clean_strava_data
```

```{r}
clean_strava_data <- clean_strava_data %>% 
dplyr::mutate(
  zEF = (EF-mean(EF))/sd(EF),
  zTE = (TE.score-mean(TE.score))/sd(TE.score),
  SAD.FIT = 0.8*EF+0.2*zTE
)
#Calculation of our unique metric
clean_strava_data
```

```{r}
clean_strava_data <- clean_strava_data %>% 
  dplyr::select(- c(Avg.Resp, Min.Resp, Max.Resp))

clean_strava_data
```

```{r}
# Splitting the data to each runner so we can look at each one individually and compare them
hound_data <- clean_strava_data %>% 
  dplyr::filter(Runner == "Hound")

collie_data <- clean_strava_data %>% 
  dplyr::filter(Runner == "Collie")
```

```{r}
collie_data %>% 
  dplyr::filter(Activity.Type == "Trail Running") %>% 
  dplyr::select(c(Title, Distance, Time, Avg.Pace, Best.Pace))
```

Graphs start below here

# Hound

## Correlation Matrix

```{r}
cor_hound <- hound_data %>% 
  dplyr::select(c(EF, Aerobic.TE, Avg.HR, Max.HR, Avg.Cadence, Avg.Pace, Avg.Stride)) %>% 
  stats::cor(use = "complete.obs") %>% 
  corrplot::corrplot(
    method = "color",
    type = "upper",
    tl.col = "black",
    tl.srt = 45,
    order= "hclust",
    addCoef.col = "black"
    )

cor_hound
```

Hound's correlation:

-   There is a strong negative correlation between EF and Avg. Pace of -0.91, which suggests that EF is tracking performance very well

-   Strong positive correlation between EF and Cadence of 0.61 and EF and Stride of 0.60, suggesting a higher performance when the running basics are also improving.

Observations from the Treadmill Running are very clear and present improvements over time

## Efficiency Factor Distribution

```{r}
hound_data %>% 
  ggplot2::ggplot(ggplot2::aes(x = EF)) +
  ggplot2::geom_histogram(
    ggplot2::aes(y = ggplot2::after_stat(density)),
    bins = ceiling((max(hound_data$EF) - min(hound_data$EF)) / ( 2 * IQR(hound_data$EF) / length(hound_data$EF)^(1/3))), # Using Freedman-Diaconis's Rule
    col = "midnightblue",
    fill = "midnightblue"
  ) +
  ggplot2::geom_vline(
    ggplot2::aes(xintercept = mean(EF, na.rm = TRUE)),
    color = "red",
    linetype = "longdash",
    size = 0.8
  ) +
  ggplot2::stat_function(
    fun = dnorm,
    args = c(mean = mean(hound_data$EF, na.rm = TRUE), sd = sd(hound_data$EF, na.rm = TRUE)),
    col = "red"
  ) +
  ggplot2::geom_rug(sides = "b") +
  ggplot2::labs(
    title = "Hound's Efficiency Factor Distribution",
    x = "Efficiency Factor",
    y = "Density"
  ) +
  ggplot2::theme(plot.title = ggplot2::element_text(hjust = 0.5))
```

It seems like Hound's performance is normally distributed around an average Efficiency Factor of 0.064. The normal distribution of the Efficiency Factor suggests that Hound has approximately as many bad days asgood days, but most days his performance is in the middle at 0.064.

```{r}
te_dist_hound <- hound_data %>% 
  ggplot2::ggplot(ggplot2::aes(x = Aerobic.TE)) +
  ggplot2::geom_histogram(
    ggplot2::aes(y = ggplot2::after_stat(density)),
    bins = ceiling((max(hound_data$Aerobic.TE) - min(hound_data$Aerobic.TE)) / ( 2 * IQR(hound_data$Aerobic.TE) / length(hound_data$Aerobic.TE)^(1/3))), # Using Freedman-Diaconis's Rule
    col = "midnightblue",
    fill = "midnightblue"
  ) +
  ggplot2::geom_vline(
    ggplot2::aes(xintercept = mean(Aerobic.TE, na.rm = TRUE)),
    color = "red",
    linetype = "longdash",
    size = 0.8
  ) +
  ggplot2::stat_function(
    fun = dnorm,
    args = c(mean = mean(hound_data$Aerobic.TE, na.rm = TRUE), sd = sd(hound_data$Aerobic.TE, na.rm = TRUE)),
    col = "red"
  ) +
  ggplot2::geom_rug(sides = "b") +
  ggplot2::labs(
    title = "Hound's Training Effect Distribution",
    x = "Training Effect",
    y = "Density"
  ) +
  ggplot2::theme(plot.title = ggplot2::element_text(hjust = 0.5))

te_dist_hound
```

```{r}
hound_data %>% 
  dplyr::select(Date, Runner, Aerobic.TE) %>%
  dplyr::group_by(Aerobic.TE) %>% 
  dplyr::count(Aerobic.TE) %>% 
  dplyr::mutate(proportion = n / nrow(hound_data)) %>% 
  ggplot2::ggplot(ggplot2::aes(x = Aerobic.TE, y = proportion)) +
  ggplot2::geom_col(fill = "midnightblue") +
  ggplot2::labs(
    title = "Hound's Proportional Training Effect",
    x = "Training Effect Zone",
    y = "Proportion (%)"
  ) +
  ggplot2::theme(plot.title = element_text(hjust = 0.5)) +
  ggplot2::scale_y_continuous(labels = scales::percent_format())
```

```{r}
hound_roll <- hound_data %>%
  dplyr::arrange(Date) %>%
  dplyr::mutate(
    SAD20 = slider::slide_dbl(
      .x = SAD.FIT,
      .f = mean,
      .before = 20,
      .after = 0,
      .complete = FALSE
    ),
    SAD200 = slider::slide_dbl(
      .x = SAD.FIT,
      .f = mean,
      .before = 200,
      .after = 0,
      .complete = FALSE
    )
  )

hound_roll %>%
  ggplot2::ggplot(ggplot2::aes(x = Date, y = SAD20)) +
  ggplot2::geom_line(col="red") +
  ggplot2::geom_line(ggplot2::aes(y = SAD200), col="blue")
```

# Collie

```{r}
collie_data %>% 
  dplyr::select(Date, Activity.Type, Distance, Calories, Time, Aerobic.TE) %>% 
  dplyr::group_by(Activity.Type) %>% 
  ggplot2::ggplot(ggplot2::aes(x = Date, y = Aerobic.TE, col = Activity.Type)) + 
  ggplot2::geom_line(ggplot2::aes(color = Activity.Type), size = 1, alpha = 0.8) +
  ggplot2::facet_wrap(~Activity.Type, scales = "free_y", ncol = 2) +
  ggplot2::labs(
    title = "Collie's Training Effect by Activity",
    x = "Date",
    y = "Aerobic Training Effect"
  ) +
  ggplot2::theme(plot.title = ggplot2::element_text(hjust = 0.5))
```

## Correlation Matrix

```{r}
cor_collie <- collie_data %>% 
  dplyr::select(c(EF, Aerobic.TE, Avg.HR, Max.HR, Avg.Cadence, Avg.Pace, Avg.Stride)) %>% 
  stats::cor(use = "complete.obs") %>% 
  corrplot::corrplot(
    method = "color",
    type = "upper",
    tl.col = "black",
    tl.srt = 45,
    order= "hclust",
    addCoef.col = "black"
    )

cor_collie
```

Note that for Collie:

-   The correlation between EF and Cadence of 0.88 and EF and Stride of 0.86 is a strong positive correlation that suggests an improvement in performance when the running basics are also improving

-   The correlation between Avg. HR and TE is 0.62, suggesting of a strong positive relationship.

-   As we expected the correlation between Avg. HR and Avg. Pace is somewhat strong but inverted; this makes sense because when as the individual runs faster its pace value decreases but heart rate increases.

    Collie’s correlations between Efficiency Factor and both cadence (0.88) and stride length (0.86) are substantially stronger than Hound’s (0.61 and 0.60). The correlation strength suggests that Collie’s efficiency is more strongly associated with consistent running mechanics, which may help explain why Collie’s overall EF is higher

## Efficiency Factor Distribution

```{r}
collie_data %>% 
  ggplot2::ggplot(ggplot2::aes(x = EF)) +
  ggplot2::geom_histogram(
    ggplot2::aes(y = ggplot2::after_stat(density)),
    bins = ceiling((max(collie_data$EF) - min(collie_data$EF)) / ( 2 * IQR(collie_data$EF) / length(collie_data$EF)^(1/3))), # Using Freedman-Diaconis's Rule
    col = "midnightblue",
    fill = "midnightblue"
  ) +
  ggplot2::geom_vline(
    ggplot2::aes(xintercept = mean(EF, na.rm = TRUE)),
    color = "red",
    linetype = "longdash",
    size = 0.8
  ) +
  ggplot2::stat_function(
    fun = dnorm,
    args = c(mean = mean(collie_data$EF, na.rm = TRUE), sd = sd(collie_data$EF, na.rm = TRUE)),
    col = "red"
  ) +
  ggplot2::geom_rug(sides = "b") +
  ggplot2::labs(
    title = "Collie's Efficiency Factor Distribution",
    x = "Efficiency Factor",
    y = "Density"
  ) +
  ggplot2::theme(plot.title = ggplot2::element_text(hjust = 0.5))
```

```{r}
cat("Mean:", mean(collie_data$EF), "SD:",sd(collie_data$EF))
```

It seems Collie's performance distribution according to the Efficiency Factor is around an average of 0.081. The left-skewed distribution of the Efficiency Factor suggests that Collie is a high-performing runner, as he is consistently performing around his average of 0.081.

```{r}
collie_data %>% 
  ggplot2::ggplot(ggplot2::aes(x = Aerobic.TE)) +
  ggplot2::geom_histogram(
    ggplot2::aes(y = ggplot2::after_stat(density)),
    bins = ceiling((max(collie_data$Aerobic.TE) - min(collie_data$Aerobic.TE)) / ( 2 * IQR(collie_data$Aerobic.TE) / length(collie_data$Aerobic.TE)^(1/3))), # Using Freedman-Diaconis's Rule
    col = "midnightblue",
    fill = "midnightblue"
  ) +
  ggplot2::geom_vline(
    ggplot2::aes(xintercept = mean(Aerobic.TE, na.rm = TRUE)),
    color = "red",
    linetype = "longdash",
    size = 0.8
  ) +
  ggplot2::stat_function(
    fun = dnorm,
    args = c(mean = mean(collie_data$Aerobic.TE, na.rm = TRUE), sd = sd(collie_data$Aerobic.TE, na.rm = TRUE)),
    col = "red"
  ) +
  ggplot2::geom_rug(sides = "b") +
  ggplot2::labs(
    title = "Collie's Training Effect Distribution",
    x = "Training Effect",
    y = "Density"
  ) +
  ggplot2::theme(plot.title = ggplot2::element_text(hjust = 0.5))
```

```{r}
collie_roll <- collie_data %>%
  dplyr::arrange(Date) %>%
  dplyr::mutate(
    SAD20 = slider::slide_dbl(
      .x = SAD.FIT,
      .f = mean,
      .before = 20,
      .after = 0,
      .complete = FALSE
    ),
    SAD200 = slider::slide_dbl(
      .x = SAD.FIT,
      .f = mean,
      .before = 200,
      .after = 0,
      .complete = FALSE
    )
  )

collie_roll %>%
  ggplot2::ggplot(ggplot2::aes(x = Date, y = SAD20)) +
  ggplot2::geom_line(col="red") +
  ggplot2::geom_line(ggplot2::aes(y = SAD200), col="blue")
```

# The Task

1.  Explain how you define fitness.

2.  Tell me about which individual is fitter?

3.  Tell me about which individual has improved more in fitness?

4.  Make a clear and concise recommendation on which individual you would want to coach. There is no `it depends here`, you have to make a decision.

Write a bullet point summary on your findings at the beginning of the document.

You must use a quarto document (.qmd) and render it into `html`.

# Metrics Meaning

-   Decompression: This means to use running as a personal wellness tool to decompress from stress

-   Max.Resp: Max.Resp = Maximum respiratory = Max rate of VO2

-   Min.Resp: Min.Resp = Minimum respiratory = Min rate of VO2

-   Min.Temp: The weather's lowest temperature during the exercise

-   <u>Stride Length:</u>

    "The distance you cover with each step of your run" [@Leicht_2023]. Additionally, @Vasudevan_2025 suggests that knowing a person's stride length can help detect a tendency to overstride, which can help prevent injuries, such as knee injuries.

-   <u>Cadence:</u>

    "The number of steps you take in one minute" [@Leicht_2023]. Furthermore, @Nike_2026 explains that cadence is important because it can be used to optimize an athlete's ground contact time, explain abnormal stride length, and running inefficiency. For instance, a lower cadence often means longer contact time with the ground, which in turn means a longer stride length, and that could mean pushing the foot too far ahead of the athlete's body; doing so is what causes knee injuries and results in wasted energy, or, in other words, loss of performance [@Nike_2026].

# What is Fitness?

According to @Cambridge_nd, the definition of fitness is "the condition of being physically strong and healthy," and @Healthlinkbc_2022 expands on that definition by adding the ability to perform physical activity. Therefore, to understand what makes an individual fit, we need to understand what enables an individual to perform physical activity and which metrics we can use to track such performance. Since our analysis focuses on fitness in running, we will focus on running performance and the specific metrics used to track it.

**The performance metrics that we will use are the following:**

-   <u>Aerobic Training Effect:</u>

    According to @Zhao_2019, Training Effect measures how each training session affects VO2 Max (Maximum Volume of Oxygen) improvement. Furthermore, @Harvard_2024 suggests that VO2 Max can indicate an individual's fitness because a high VO2 Max correlates with better physical fitness and is also associated with "a lower risk for cardiovascular disease and increased longevity." The reason VO2 Max correlates with better physical fitness and health is that it measures the amount of oxygen an athlete consumes during exercise [@Harvard_2024]. Oxygen consumption matters because, for a muscle to generate energy as ATP, it needs oxygen; therefore, the more oxygen the athlete can breathe in, the more energy the muscle can produce to perform the exercise [@Harvard_2024].

    @Zhao_2019 suggests the following description for the Training Effect ranges: (Turn this to a GT table later)

    -   0.0 - 0.9: No aerobic effect

    -   1.0 - 1.9: Minor aerobic effect

    -   2.0 - 2.9: Maintaining aerobic effect

    -   3.0 - 3.9: Improving aerobic effect

    -   4.0 - 4.9: Highly improving aerobic effect

    -   5.0: Overreaching aerobic effect

-   <u>Efficiency Factor:</u>

    According to @nxtrun_2025, the efficiency factor (EF) is a performance measure that quantifies the speed an individual achieves per heartbeat. The reason EF matters is that an efficient individual uses less oxygen, thereby requiring less energy from the heart to pump oxygen to the muscles to maintain a given speed.

    @nxtrun_2025 and @Friel suggest that the efficiency factor calculation is the following:

    $$ EF = \frac{\text{Speed}}{\text{Heart Rate}} $$

    Since we have an average pace, we can use it to find the average speed, since pace and speed are inverses. That means that:

    $$ \text{Speed} = \frac{1}{\text{Pace}} $$

It is important to note that Training Effect is a short-term performance measure since it only tracks the Max VO2 improvement from a single workout, the last exercise the individual performed. In contrast, the Efficiency Factor is useful for long-term performance tracking because it relates speed to heart rate, allowing us to observe improvements in running efficiency over time.

To determine which individual is the most fit, we will use the current performance on their metrics. However, to determine which individual we will train, we will estimate each individual's potential by forecasting their future metrics and comparing them to select the one we want to train.

In order to determine who is more fit we will be primarily using Efficiency Factor. Aerobic Training Effect will be a more important effect in our analysis of rate of improvement.

As we saw the importance of both factors in determining who to train we used these factors to create our own metric which we have named the "SAD" metric. The name comes from the initials of each of our names. This metric hopes to combine the strengths of both Aerobic TE and Efficiency Factor to make a definitive choice of which runner we will train.

# Comparison of Aerobic Training Effect between the runners:

```{r}
hound_data %>% 
  dplyr::select(Date, Activity.Type, Distance, Calories, Time, Aerobic.TE) %>% 
  dplyr::group_by(Activity.Type) %>% 
  ggplot2::ggplot(ggplot2::aes(x = Date, y = Aerobic.TE, col = Activity.Type)) + 
  ggplot2::geom_line(ggplot2::aes(color = Activity.Type), size = 1, alpha = 0.8) +
  ggplot2::facet_wrap(~Activity.Type, scales = "free_y", ncol = 2) +
  ggplot2::labs(
    title = "Hound's Training Effect by Activity",
    x = "Date",
    y = "Aerobic Training Effect"
  ) +
  ggplot2::theme(plot.title = ggplot2::element_text(hjust = 0.5))
```

```{r}
collie_data %>% 
  dplyr::select(Date, Activity.Type, Distance, Calories, Time, Aerobic.TE) %>% 
  dplyr::group_by(Activity.Type) %>% 
  ggplot2::ggplot(ggplot2::aes(x = Date, y = Aerobic.TE, col = Activity.Type)) + 
  ggplot2::geom_line(ggplot2::aes(color = Activity.Type), size = 1, alpha = 0.8) +
  ggplot2::facet_wrap(~Activity.Type, scales = "free_y", ncol = 2) +
  ggplot2::labs(
    title = "Collie's Training Effect by Activity",
    x = "Date",
    y = "Aerobic Training Effect"
  ) +
  ggplot2::theme(plot.title = ggplot2::element_text(hjust = 0.5))
```

```{r}
#graphs for TE
collie_data %>% 
  dplyr::select(Date, Runner, Aerobic.TE) %>%
  dplyr::group_by(Aerobic.TE) %>% 
  dplyr::count(Aerobic.TE) %>% 
  dplyr::mutate(proportion = n / nrow(collie_data)) %>% 
  ggplot2::ggplot(ggplot2::aes(x = Aerobic.TE, y = proportion)) +
  ggplot2::geom_col(fill = "midnightblue") +
  ggplot2::labs(
    title = "Collie's Proportional Training Effect",
    x = "Training Effect Zone",
    y = "Proportion (%)"
  ) +
  ggplot2::theme(plot.title = element_text(hjust = 0.5)) +
  ggplot2::scale_y_continuous(labels = scales::percent_format())
```

```{r}
hound_data %>% 
  dplyr::select(Date, Activity.Type, Distance, Calories, Time, Aerobic.TE) %>% 
  dplyr::group_by(Activity.Type) %>% 
  dplyr::summarise(Aerobic.TE = mean(Aerobic.TE)) %>% 
  ggplot2::ggplot(ggplot2::aes(x = Activity.Type, y = Aerobic.TE, fill = Activity.Type)) + 
  ggplot2::geom_col()
```

```{r}
hound_data %>% 
  dplyr::select(Date, Runner, Aerobic.TE) %>%
  dplyr::group_by(Aerobic.TE) %>% 
  dplyr::count(Aerobic.TE) %>% 
  dplyr::mutate(proportion = n / nrow(hound_data)) %>% 
  ggplot2::ggplot(ggplot2::aes(x = Aerobic.TE, y = proportion)) +
  ggplot2::geom_col(fill = "midnightblue") +
  ggplot2::labs(
    title = "Hound's Proportional Training Effect",
    x = "Training Effect Zone",
    y = "Proportion (%)"
  ) +
  ggplot2::theme(plot.title = element_text(hjust = 0.5)) +
  ggplot2::scale_y_continuous(labels = scales::percent_format())
```

```{r}
collie_data %>% 
  ggplot2::ggplot(ggplot2::aes(x = Aerobic.TE)) +
  ggplot2::geom_histogram(
    ggplot2::aes(y = ggplot2::after_stat(density)),
    bins = ceiling((max(collie_data$Aerobic.TE) - min(collie_data$Aerobic.TE)) / ( 2 * IQR(collie_data$Aerobic.TE) / length(collie_data$Aerobic.TE)^(1/3))), # Using Freedman-Diaconis's Rule
    col = "midnightblue",
    fill = "midnightblue"
  ) +
  ggplot2::geom_vline(
    ggplot2::aes(xintercept = mean(Aerobic.TE, na.rm = TRUE)),
    color = "red",
    linetype = "longdash",
    size = 0.8
  ) +
  ggplot2::stat_function(
    fun = dnorm,
    args = c(mean = mean(collie_data$Aerobic.TE, na.rm = TRUE), sd = sd(collie_data$Aerobic.TE, na.rm = TRUE)),
    col = "red"
  ) +
  ggplot2::geom_rug(sides = "b") +
  ggplot2::labs(
    title = "Collie's Training Effect Distribution",
    x = "Training Effect",
    y = "Density"
  ) +
  ggplot2::theme(plot.title = ggplot2::element_text(hjust = 0.5))
```

```{r}
collie_data %>% 
  dplyr::select(Date, Runner, Aerobic.TE) %>%
  dplyr::group_by(Aerobic.TE) %>% 
  dplyr::count(Aerobic.TE) %>% 
  dplyr::mutate(proportion = n / nrow(collie_data)) %>% 
  ggplot2::ggplot(ggplot2::aes(x = Aerobic.TE, y = proportion)) +
  ggplot2::geom_col(fill = "midnightblue") +
  ggplot2::labs(
    title = "Collie's Proportional Training Effect",
    x = "Training Effect Zone",
    y = "Proportion (%)"
  ) +
  ggplot2::theme(plot.title = element_text(hjust = 0.5)) +
  ggplot2::scale_y_continuous(labels = scales::percent_format())
```

(GRAPH for comparison goes here)

(commentary on graphs for each runner)

(either second set of charts if they are more insightful here or conclusion for AE)

# Comparison of Efficiency Factor between the runners:

```{r}
hound_data %>% 
  ggplot2::ggplot(ggplot2::aes(x = EF)) +
  ggplot2::geom_histogram(
    ggplot2::aes(y = ggplot2::after_stat(density)),
    bins = ceiling((max(hound_data$EF) - min(hound_data$EF)) / ( 2 * IQR(hound_data$EF) / length(hound_data$EF)^(1/3))), # Using Freedman-Diaconis's Rule
    col = "midnightblue",
    fill = "midnightblue"
  ) +
  ggplot2::geom_vline(
    ggplot2::aes(xintercept = mean(EF, na.rm = TRUE)),
    color = "red",
    linetype = "longdash",
    size = 0.8
  ) +
  ggplot2::stat_function(
    fun = dnorm,
    args = c(mean = mean(hound_data$EF, na.rm = TRUE), sd = sd(hound_data$EF, na.rm = TRUE)),
    col = "red"
  ) +
  ggplot2::geom_rug(sides = "b") +
  ggplot2::labs(
    title = "Hound's Efficiency Factor Distribution",
    x = "Efficiency Factor",
    y = "Density"
  ) +
  ggplot2::theme(plot.title = ggplot2::element_text(hjust = 0.5))
```

```{r}
cat("Mean:", mean(hound_data$EF), "SD:",sd(hound_data$EF))
```

It seems like Hound's performance is normally distributed around an average Efficiency Factor of 0.064. The normal distribution of the Efficiency Factor suggests that Hound has approximately as many bad days asgood days, but most days his performance is in the middle at 0.064.

```{r}
collie_data %>% 
  ggplot2::ggplot(ggplot2::aes(x = EF)) +
  ggplot2::geom_histogram(
    ggplot2::aes(y = ggplot2::after_stat(density)),
    bins = ceiling((max(collie_data$EF) - min(collie_data$EF)) / ( 2 * IQR(collie_data$EF) / length(collie_data$EF)^(1/3))), # Using Freedman-Diaconis's Rule
    col = "midnightblue",
    fill = "midnightblue"
  ) +
  ggplot2::geom_vline(
    ggplot2::aes(xintercept = mean(EF, na.rm = TRUE)),
    color = "red",
    linetype = "longdash",
    size = 0.8
  ) +
  ggplot2::stat_function(
    fun = dnorm,
    args = c(mean = mean(collie_data$EF, na.rm = TRUE), sd = sd(collie_data$EF, na.rm = TRUE)),
    col = "red"
  ) +
  ggplot2::geom_rug(sides = "b") +
  ggplot2::labs(
    title = "Collie's Efficiency Factor Distribution",
    x = "Efficiency Factor",
    y = "Density"
  ) +
  ggplot2::theme(plot.title = ggplot2::element_text(hjust = 0.5))
```

```{r}
cat("Mean:", mean(collie_data$EF), "SD:",sd(collie_data$EF))
```

```{r}
 hound_data %>% 
  dplyr::select(Date, Avg.Pace, Distance, Avg.HR, Activity.Type, EF) %>%
  ggplot2::ggplot(ggplot2::aes(x = Date, y = EF)) +  
  ggplot2::geom_line(ggplot2::aes(color = Activity.Type), size = 1, alpha = 0.8)+
  ggplot2::facet_wrap(~Activity.Type, scales = "free_y", ncol = 2) +
  ggplot2::labs(
    title = "Hound's Efficiency Factor by Activity",
    x = "Date",
    y = "Efficiency Factor") +
  ggplot2::theme(plot.title = ggplot2::element_text(hjust = 0.5))
```

```{r}
 collie_data %>% 
  dplyr::select(Date, Avg.Pace, Distance, Avg.HR, Activity.Type, EF) %>%
  ggplot2::ggplot(ggplot2::aes(x = Date, y = EF)) +  
  ggplot2::geom_line(ggplot2::aes(color = Activity.Type), size = 1, alpha = 0.8)+
  ggplot2::facet_wrap(~Activity.Type, scales = "free_y", ncol = 2) +
  ggplot2::labs(
    title = "Collie's Efficiency Factor by Activity",
    x = "Date",
    y = "Efficiency (EF)")+
     ggplot2::theme(plot.title = ggplot2::element_text(hjust = 0.5))
 
```

```{r}
hound_data %>% 
  dplyr::select(Date, Activity.Type, Distance, Calories, Time, EF) %>% 
  dplyr::group_by(Activity.Type) %>% 
  dplyr::summarise(EF = mean(EF)) %>% 
  ggplot2::ggplot(ggplot2::aes(x = Activity.Type, y = EF, fill = Activity.Type)) + 
  ggplot2::geom_col()
```

```{r}
collie_data %>% 
  dplyr::select(Date, Activity.Type, Distance, Calories, Time, EF) %>% 
  dplyr::group_by(Activity.Type) %>% 
  dplyr::summarise(EF = mean(EF)) %>% 
  ggplot2::ggplot(ggplot2::aes(x = Activity.Type, y = EF, fill = Activity.Type)) + 
  ggplot2::geom_col()
```

(GRAPH for comparison goes here)

(commentary on graphs for each runner)

(either second set of charts if they are more insightful here or conclusion for EF)

# Our thoughts on the use of each metric for determining who is more fit:

Each metric captures a different aspect of running performance, so no single measure should be treated as a complete definition of fitness. Efficiency Factor (EF) is useful because it reflects how much speed is being produced for a given heart rate, making it a practical indicator of cardiovascular efficiency. Training Effect (TE) is useful for describing the physiological impact of a run, but on its own it reflects session stimulus more than underlying fitness. The custom combined metric is useful because it incorporates both efficiency and training stimulus, giving a broader session-quality measure than either variable alone. The moving averages add further value by smoothing short-term fluctuations and making the metric easier to interpret over time. Taken together, these measures are better viewed as complementary indicators rather than standalone tests of fitness.

Summary of each metric here, how impactful each is for determining fitness.

We may want to use out metric for fitness here or we might want to use that metric for who we want to train.

We say here who is more fit by out definition.

# Our thoughts on the use of each metric for determining who has improved more:

For assessing improvement, trend-based measures are more useful than single-run values. Raw EF and TE from one session may be influenced by daily conditions such as fatigue, route, weather, or workout intensity, so they are limited as direct measures of improvement on their own. The custom combined metric is more helpful because it summarizes both efficiency and training quality in one score, but its interpretation is stronger when viewed across multiple runs rather than in isolation. For this reason, the moving averages are especially useful for evaluating improvement, since they smooth out short-term noise and highlight the direction of the metric over time. This makes them more appropriate for comparing change in running performance across longer periods.

# Our recommendation on who we want to coach:

The fitness metric we came up with is a custom performance measure that is designed to combine efficiency and aerobic training effect into a single score.

The first component is Efficiency Factor (EF), which captures how much speed is being produced relative to heart rate. In this case, pace is converted into speed, and EF is calculated as speed divided by average heart rate. A higher EF suggests the runner is moving faster for a given cardiovascular effort, which can be interpreted as better efficiency.

The second component is Aerobic Training Effect (TE), which reflects the physiological impact of the run. Based on the TE interpretation ranges already provided in the document, TE is converted into a custom score that increases as the aerobic effect becomes more meaningful, with the highest reward assigned to strongly improving aerobic effect levels. A TE of 5 is assigned a score of 0, since it represents an overreaching effect and is treated in this metric as excessively fatiguing rather than ideal for regular training. This scoring structure is intended to reward productive aerobic stimulus without automatically rewarding maximal effort. Since, EF and the TE score are measured on different scales, they are first scaled before being combined.

The final metric is therefore intended to summarize both how efficiently the run was performed and how productive the training stimulus, it is weighted using an 80/20 ratio, with 80% assigned to EF and 20% assigned to the TE score. The goal of this metric is not to create a perfect universal measure of performance, but to build a practical session-quality indicator that can be tracked over time to identify whether recent running performance and training quality are improving, declining, or remaining stable.

To examine the custom metric over time, moving averages based on 20 and 200 observations were computed. The 20-observation moving average provides a short-run view of the metric and responds more quickly to recent fluctuations, whereas the 200-observation moving average provides a smoother long-run representation of the series. Plotting both moving averages helps reduce short-term noise and makes the time-series behavior of the metric easier to visualize.

Our recommendation is to coach Collie. Based on the moving-average graphs, Collie shows the stronger upward pattern in the custom metric over time, especially in the longer-run 200-observation average, which gradually rises across the sample. The shorter-run 20-observation average also spends much of the later period above the longer-run average, indicating stronger recent performance relative to Collie’s own longer-term trend. In contrast, Hound’s 200-observation average trends downward over time, and the 20-observation average is more volatile, with larger swings above and below the longer-run line. Taken together, the graphs suggest that Collie provides the clearer case for coaching support because the recent trajectory of the metric is more favorable and shows stronger momentum in the underlying performance trend.

# Notes

-   The distance for everything is in Km, but Track Running is in meters.

-   Collie doesn't have any data under Track Running, but I noticed that under Title in Trail Running there's at least 1 observation with a Title description of Track Running - it's on the last page.

-   Collie never ran Street Running or at least he doesn't have any records of running one.

-   Observations from the Treadmill Running are very clear and present improvements over time

# Assumptions

-   We are assuming that Collie mistakenly entered the "Track Run" as the title for one of his Trail Running data, primarly because he does not have any data on Track Running so we do not expect him to have only one.
